---
title: "STAT 5443-Computational Statistics-Midterm hw"
author: "Md Shahbaz Alam"
date: "2023-09-28"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list = ls())
current_path <- rstudioapi::getSourceEditorContext()$path
setwd(dirname(current_path))
library(tidyverse)
set.seed(729)
```

\newpage
## Question 2: Inverse transform Sampling

```{r}
#------------- Problem 2 (2): Inverse Transform Sampling
# Defining the inverse CDF: 2 * u^(1/4)
inverse_cdf <- function(u) {
  return(2 * u ^(1/4))
}

# Generate 10,000 random samples using ITS
set.seed(729) 
n_samples <- 10000
u <- runif(n = n_samples, 0, 1)
random_samples <- inverse_cdf(u)
hist(random_samples,probability = T)
curve(expr = .25*(x^3),from = 0,to = 2,add = T, col="dodgerblue4")
```


<!-- ```{r echo = FALSE, include=TRUE} -->
<!-- linedf <- data.frame(x = seq(from = 0, to = 2, length.out = 100)) %>%  -->
<!--   mutate(pdf = .25 * x^3) -->
<!-- # after_stat(density) -->
<!-- p2 <- ggplot() + -->
<!--   geom_histogram(aes(x = random_samples, y = after_stat(density)), fill = "white", col = "#003153") +  -->
<!--   geom_line(data = linedf, aes(x = x, y = pdf), color = "dodgerblue4", lwd = 1.1) + -->
<!--   scale_x_continuous(breaks = seq(0, 2, by = .5)) + -->
<!--   # geom_density(aes( x = rnorm(5000, 20, 10000)), col = "seagreen", lwd = 1) + -->
<!--   labs(title = expression(paste("Histogram of Samples using Inverse Transform Sampling with Density Line for : ", f(x) == frac(1, 4) ~ x^3)), -->
<!--        x = "", -->
<!--        y = "Density", color = "Blue") + -->
<!--   theme_classic() +  -->
<!--   theme( -->
<!--     # plot.title = element_markdown(lineheight = .75, size= 9, face="bold", hjust = 0.5), -->
<!--     plot.title = element_text(size = 9, face = "bold", color = 'black', hjust = 0.5), -->
<!--     axis.title = element_text(size = 9, face="bold"), -->
<!--     legend.title = element_blank(), legend.position = "top", -->
<!--     panel.border = element_rect(color = "black", fill = NA, linewidth = 1) -->
<!--   ) -->
<!-- print(p2) -->
<!-- ``` -->


\newpage
## Question 3:

The marginal density X and Y are symmetric function, so their density looks similar.
```{r}
marginal_pdf <- function(v) {
  
  .25 * (v + (3/(2*sqrt(2))) * sqrt(v) )
}
curve(marginal_pdf, from = 0, to = 2, col="dodgerblue4")
```


<!-- ```{r echo=FALSE, include=TRUE} -->

<!-- marginal_density <- data.frame(x = seq(from = 0, to = 2, length.out = 1000)) %>%  -->
<!--   mutate(pdf = marginal_pdf(x)) -->
<!-- # after_stat(density) -->
<!-- marginal_plot3 <- ggplot() + -->
<!--   geom_line(data = marginal_density, aes(x = x, y = pdf), color = "dodgerblue4", lwd = 1.1) + -->
<!--   # geom_density(aes( x = rnorm(5000, 20, 10000)), col = "seagreen", lwd = 1) + -->
<!--   labs(title = "Marginal Density of X or Y , Both Density are Similar", -->
<!--        x = "", -->
<!--        y = "Marginal Density", color = "Blue") + -->
<!--   theme_classic() +  -->
<!--   theme( -->
<!--     # plot.title = element_markdown(lineheight = .75, size= 9, face="bold", hjust = 0.5), -->
<!--     plot.title = element_text(size = 9, face = "bold", color = 'black', hjust = 0.5), -->
<!--     axis.title = element_text(size = 9, face="bold"), -->
<!--     legend.title = element_blank(), legend.position = "top", -->
<!--     panel.border = element_rect(color = "black", fill = NA, linewidth = 1) -->
<!--   ) -->
<!-- print(marginal_plot3) -->
<!-- ``` -->


\newpage
## Gibss Sampler:

```{r echo = TRUE}
set.seed(729)
conditional_distribution <- function(x, y) {
  c_x <- (3 / (4 * sqrt(2))) * (y + (3 / (2 * sqrt(2))) * sqrt(y))^-1
  return(c_x*(y * sqrt(x) + x * sqrt(y)))
}

rejection_sampling <- function(y) {
  
  generate = TRUE
  
  while(generate) {
    
    z <- runif(1, min = 0,max = 2)
    u <- runif(1, min = 0, max = 1)
    
    if(u < conditional_distribution(x = z, y = y))
      generate = FALSE
    
  }
  return(z)
}

# Gibbs sampling
x = y = 1

for(i in 2:10000) {
  x[i] = rejection_sampling(y[i-1])
  y[i] = rejection_sampling(x[i])
}
```

\newpage
As we can see the pdf of marginal X and the histogram of X using the Gibbs sampling
has matched. Thus we can say, the algorithm works well.
```{r}
hist(x, probability = T)
curve(marginal_pdf,from = 0,to = 2, add = T, col="dodgerblue4")
```

<!-- ```{r echo = FALSE, include=TRUE} -->
<!-- # marginal_density <- data.frame(x = seq(from = 0, to = 2, length.out = 1000)) %>%  -->
<!-- #   mutate(pdf = marginal_pdf(x)) -->
<!-- # after_stat(density) -->
<!-- x_gibbs <- ggplot() + -->
<!--   geom_histogram(aes(x = x, y = after_stat(density)), bins = 30, fill = "white", col = "#003153") +  -->
<!--   geom_line(data = marginal_density, aes(x = x, y = pdf), color = "dodgerblue4", lwd = 1.1) + -->
<!--   # geom_density(aes( x = rnorm(5000, 20, 10000)), col = "seagreen", lwd = 1) + -->
<!--   labs(title = "Histogram of X using Gibbs Sampling with Marginal Density Line", -->
<!--        x = "Gibbs Sample of X", -->
<!--        y = "Density", color = "Blue") + -->
<!--   theme_classic() +  -->
<!--   theme( -->
<!--     # plot.title = element_markdown(lineheight = .75, size= 9, face="bold", hjust = 0.5), -->
<!--     plot.title = element_text(size = 9, face = "bold", color = 'black', hjust = 0.5), -->
<!--     axis.title = element_text(size = 9, face="bold"), -->
<!--     legend.title = element_blank(), legend.position = "top", -->
<!--     panel.border = element_rect(color = "black", fill = NA, linewidth = 1) -->
<!--   ) -->
<!-- print(x_gibbs) -->

<!-- ``` -->

\newpage
As we can see the pdf of marginal Y and the histogram of Y using the Gibbs sampling 
has matched. Thus we can say, the algorithm works well.

```{r}
hist(y, probability = T)
curve(marginal_pdf, from = 0, to = 2, add = T, col="dodgerblue4")
```

<!-- ```{r echo = FALSE, include=TRUE} -->
<!-- # marginal_density <- data.frame(x = seq(from = 0, to = 2, length.out = 1000)) %>%  -->
<!-- #   mutate(pdf = marginal_pdf(x)) -->
<!-- # after_stat(density) -->
<!-- y_gibbs <- ggplot() + -->
<!--   geom_histogram(aes(x = y, y = after_stat(density)), bins = 30, fill = "white", col = "#003153") +  -->
<!--   geom_line(data = marginal_density, aes(x = x, y = pdf), color = "dodgerblue4", lwd = 1.1) + -->
<!--   labs(title = "Histogram of Y using Gibbs Sampling with Marginal Density Line", -->
<!--        x = "Gibbs Sample of Y", -->
<!--        y = "Density", color = "Blue") + -->
<!--   theme_classic() +  -->
<!--   theme( -->
<!--     plot.title = element_text(size = 9, face = "bold", color = 'black', hjust = 0.5), -->
<!--     axis.title = element_text(size = 9, face="bold"), -->
<!--     legend.title = element_blank(), legend.position = "top", -->
<!--     panel.border = element_rect(color = "black", fill = NA, linewidth = 1) -->
<!--   ) -->
<!-- print(y_gibbs) -->

<!-- ``` -->


\newpage
```{r}
x <- c(1.5, 2.6, 1.0, 3.0, 2.8)
n <- length(x)
ss <- sum((x - 2)^2)
n_samples <- 10000
ig_sample <- 1/rgamma(n_samples, shape = n/2, rate = ss/2)
```


```{r echo = FALSE, include=TRUE}
ig_histogram <- ggplot() +
  geom_histogram(aes(x = ig_sample, y = after_stat(density)), bins = 80, fill = "white", col = "#003153") + 
  # geom_line(data = marginal_density, aes(x = x, y = pdf), color = "dodgerblue4", lwd = 1.1) +
  labs(title = expression(paste("Histogram of Conditional Posterior of ", sigma^2)),
       x = "Inverse Gamma Sample",
       y = "Density", color = "Blue") + 
  theme_classic() + 
  theme(
    plot.title = element_text(size = 9, face = "bold", color = 'black', hjust = 0.5),
    axis.title = element_text(size = 9, face="bold"),
    legend.title = element_blank(), legend.position = "top",
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1)
  )
print(ig_histogram)

```

