---
title: "`Simulation Exercise with Solution`"
author: "`Md Shahbaz Alam`"
date: "`r Sys.Date()`"
output: pdf_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F,comment = "  ", warning = FALSE, fig.width=13, fig.height=17)
# knitr::opts_chunk$set() 
rm(list = ls())
library(tidyverse)
library(ggpubr)
set.seed(729)
```


### Solution Q1:
*Monte-Carlo* approximation for the integral
$$\int_{0}^{\infty}\frac{\theta^6\exp^{-\theta^2 - \theta^3 - 2\theta}}{1 + 3\theta^2}d\theta$$
We aim to write this integral as $\int f(\theta) g(\theta)$ where $g(\theta)$ is a known density that is easy to simulate from. Then, we draw a large number of samples from $g(\theta)$, evaluate $f(\theta)$ for each one of those samples and take the mean, which will be the *Monte-Carlo* approximation for the above mentioned integral.
\
Approach 1: We use $Gamma(7, 2)$ as $g(\theta)$. 

Now, 
$$g(\theta) = \frac{2^7}{\Gamma(7)} \exp^{-2\theta}\theta^{7-1}$$
Thus our $f(\theta)$ will become,
$$f(\theta) = \frac{\Gamma(7)}{2^7} \frac{\exp^{-\theta^2 - \theta^3}}{1 + 3\theta^2}d\theta$$  


Approach 2: We use $exp(2)$ as $g(\theta)$. 

Now, $$g(\theta) = 2e^{-2\theta}$$
Thus our $f(\theta)$ will become,
$$f(\theta) = \frac{1}{2} \frac{\theta^6\exp^{-\theta^2 - \theta^3}}{1 + 3\theta^2}d\theta$$  


```{r include = FALSE, echo=FALSE}
# -------------- Question 1: Monte Carlo simulation to estimate an integral in two different ways:

# approach 1: Gamma(7,2) as g(x). So f(x) will be that expression

f1x <- function(x){
  expression <- (gamma(7)/2^7)*(exp(-x^2-x^3)/(1 + 3 * x^2))
  return(expression)
}

# approach 2: exp(2) as g(x). So f(x) will be that expression

f2x <- function(x){
  expression <- (.5 * x^6 * exp(-x^2-x^3))/ (1+3*x^2)
  return(expression)
}

# Exact integral value:
exact_val <- integrate(f = function(x) ((x^6*exp(-x^2-x^3-2*x))/(1 + 3 * x^2)), lower = 0, upper = Inf)
print(exact_val)

# Approach 1: f(x) = (gamma(7)/2^7)*(exp(-x^2-x^3)/(1 + 3 * x^2)), g(x) = Gamma(7, 2)
approach_1 <- function(num_points) {
  samples <- rgamma(num_points, shape = 7, rate = 2)
  cal_val <- f1x(samples)
  integral_approximation <- mean(cal_val)
  return(integral_approximation)
}

# Approach 2: f(x) = (.5 * x^6 * exp(-x^2-x^3))/ (1+3x^2), g(x) = exp(2)
approach_2 <- function(num_points) {
  samples <- rexp(num_points, rate = 2)
  cal_val <- f2x(samples)
  integral_approximation <- mean(cal_val)
  return(integral_approximation)
}

# number of random points to generate
num_points <- 100000

# Computing integral approximation using both approach
approximation_1 <- approach_1(num_points) # Gamma(7,2)
approximation_2 <- approach_2(num_points) # Exp(2)


```


```{r include = TRUE}

# Printing results
options(scipen=999) 
cat("Output: 
    \n\tActual result: ", round(exact_val$value, 8), 
    "\n\tMonte-Carlo approximation using Gamma(7, 2) PDF: ", round(approximation_1, 8), 
    "\n\tMonte-Carlo approximation using Exponential(2) PDF:", round(approximation_2, 8), "\n")

```

\newpage
### Solution Q2:
  
*Monte-Carlo* approximation for the integral
$$\int_{1.5}^{6}\frac{\log\theta}{2 + \theta^2}d\theta$$ 

As per procedure mentioned in *solution-1*, we use $unif(1.5, 6)$ as $g(\theta)$. Now, 
$$g(\theta) = \frac{1}{6-1.5} = \frac{1}{4.5}$$
Thus our $f(\theta)$ will become, 
$$f(\theta) = 4.5 \space\frac{\log(\theta)}{2 + \theta^2}$$

```{r include = FALSE, echo=FALSE}
# -------------- Question 2: Monte Carlo method for a integral:

# approach : Unif(1.5, 6) as g(x). So f(x) will be that following expression
fx <- function(x) {
  expression <- (4.5 * log(x))/(2 + x^2)
  return(expression)
}

# Exact integral value:
exact_val <- integrate(f = function(x) log(x)/(2 + x^2), lower = 1.5, upper = 6)
print(exact_val)

# Approach: f(x) = (4.5 * log(x))/(2 + x^2), g(x) = unif(1.5, 6)
mc_app <- function(num_points) {
  samples <- runif(num_points, min = 1.5, max = 6)
  cal_val <- fx(samples)
  integral_approximation <- mean(cal_val)
  return(integral_approximation)
}

```


```{r include = TRUE}
# Computing integral approximation
num_points <- 100000
approximate_val<- mc_app(num_points) # unif(1.5, 6)
options(scipen=999)
cat("Output: 
    \n\tActual result: ", round(exact_val$value, 8), 
    "\n\tMonte-Carlo approximation using Uniform(1.5, 6) PDF:", round(approximate_val, 8), "\n")
```


## Solution 3:
Metropolis-Hastings Simulation to draw sample where target density is:
$$\theta \sim Gamma(2,\space 4)$$

```{r include = FALSE, echo = FALSE}
# Q3: Metrpolis_hastings Algorthm
alpha <- 2; beta <- 4
target_density <- function(x){
  return(dgamma(x, shape = alpha, rate = beta))
}

true_mean <- alpha/beta
mh_alg <- function( initial_theta = 5, num_steps, burn_in, proposal_variance){
  
  set.seed(729)
  theta <- numeric(num_steps)
  theta[1] <- initial_theta
  
  accepted_proposal <- 0
  
  for(i in 2 : num_steps){
    # propose a theta value from Normal density proposal
    candidate <- rnorm(1, mean = theta[i-1], sd = sqrt(proposal_variance))
    
    # Target ratio: decides whether we accept the proposed theta as current
    acceptance_ratio <- min(1, target_density(candidate)/target_density(theta[i-1])) # markov kernel cancel out
    
    unif_value <- runif(1)
    
    if(unif_value < acceptance_ratio){
      
      # accept proposed theta
      theta[i] <- candidate
      accepted_proposal <- accepted_proposal + 1
      
    } else {
      
      # reject proposal and take immediate previous theta as current theta
      theta[i] <- theta[i - 1]
    }
    
  }
  
  theta_aburn <- theta[(burn_in + 1):num_steps] # samples after burn_in
  
  # require to decide proposal variance
  acceptance_rate <- accepted_proposal/num_steps
  
  return(list(theta = theta, theta_aburn = theta_aburn, acceptance_rate = acceptance_rate))
}

#  using proposal variance = 0.1
approach_1 <- mh_alg(num_steps = 50000, burn_in = 10000, proposal_variance = 0.1)

#  using proposal variance = 2
approach_2 <- mh_alg(num_steps = 50000, burn_in = 10000, proposal_variance = 2)

# Calculate the estimated mean for approach 1
estimated_mean_1 <- mean(approach_1$theta_aburn)

# Calculate the estimated mean for approach 2
estimated_mean_2 <- mean(approach_2$theta_aburn)


```

From target distribution True mean is 0.5

* Approach 1: Normal distribution as a proposal with variance = 0.1 and $\theta_0 = 5$

```{r include = TRUE}

# Print the results
cat("\tAcceptance Rate:", approach_1$acceptance_rate, 
    "\n\tEstimated Mean:", estimated_mean_1, 
    "\n\tDifference from True Mean:", abs(estimated_mean_1 - true_mean), "\n")

```

\newpage
* Approach 1: Normal distribution as a proposal with variance = 0.1 and $\theta_0 = 5$

```{r include=FALSE, echo=FALSE}

mh_df1 <- data.frame(theta = approach_1$theta) %>% 
  mutate(index = seq_along(1: nrow(.)),
         burn_in = if_else(index <= 10000, "Burn in", "Accepted"))

col_pal <- c("#003153", "red3")

prop_plot1 <- ggplot() + 
    geom_density(aes( x = rgamma(40000, 2, 4)), col = "#003153", lwd = 1) + 
    geom_density(aes( x = rnorm(40000, 5, sqrt(0.1))), col = "red3", lwd = 1) +
    
    labs(title = expression(paste("Actual density ", Gamma(2, 4), " vs proposal ", N(theta = 5, 0.1))),
       x = expression(theta),
       y =  "Density") +
    theme_classic() + 
    theme(
      # plot.title = element_markdown(lineheight = .75, size= 9, face="bold", hjust = 0.5),
      plot.title = element_text(size= 12, face = "bold", color = 'black', hjust = 0.5),
      axis.title = element_text(size = 9, face="bold"),
      legend.title = element_blank(),
      legend.text = element_text(size = 9, face = "bold"), legend.position = c(.95, .9),
      legend.background = element_rect(color = "black", fill= NA, size = 1, linetype= "solid"),
      panel.border = element_rect(color = "black", fill = NA, linewidth = 1)) +
      scale_color_manual(values = col_pal)



iteration_plot1 <- ggplot(data = mh_df1) + 
    geom_line(aes(x = index, y = theta, color = burn_in)) +
    labs(title = "Metropolis Hastings using Normal proposal with starting point 5",
         x = "Iteration Number",
         y = expression(theta)) +
    theme_classic() + 
    theme(
      # plot.title = element_markdown(lineheight = .75, size= 9, face="bold", hjust = 0.5),
      plot.title = element_text(size= 12, face = "bold", color = 'black', hjust = 0.5),
      axis.title = element_text(size = 9, face="bold"),
      legend.title = element_blank(),
      legend.text = element_text(size = 9, face = "bold"), legend.position = c(.95, .9),
      legend.background = element_rect(color = "black", fill= NA, size = 1, linetype= "solid"),
      panel.border = element_rect(color = "black", fill = NA, size = 1)) +
    scale_color_manual(values = col_pal)

zoomed_plot1 <-  mh_df1 %>% 
  filter(9950< index & index <10050) %>% 
  ggplot(data = .) + 
  geom_line(aes(x = index, y = theta, color = burn_in)) +
  labs(title = "100 iteration Zoomed in ",
       x = "Iteration Number",
       y = expression(theta)) +
  theme_classic() + 
    theme(
      # plot.title = element_markdown(lineheight = .75, size= 9, face="bold", hjust = 0.5),
      plot.title = element_text(size= 12, face = "bold", color = 'black', hjust = 0.5),
      axis.title = element_text(size = 9, face="bold"),
      legend.title = element_blank(),
      legend.text = element_text(size = 9, face = "bold"), legend.position = c(.95, .9),
      legend.background = element_rect(color = "black", fill= NA, size = 1, linetype= "solid"),
      panel.border = element_rect(color = "black", fill = NA, size = 1)) +
    scale_color_manual(values = col_pal)
  

plot_mh1 <- ggplot(data = filter(mh_df1, burn_in == "Accepted")) +
    geom_histogram(aes(x = theta, y = after_stat(density)), bins = 100, fill = "white", col = "#003153") + 
    geom_density(aes( x = rgamma(40000, 2, 4)), col = "seagreen", lwd = 1) +
    labs(title = expression(paste("Histogram of simulated sample using M-H algorithm from ", Gamma(2, 4)), " density"),
         x = expression(theta),
         y = "Density", color = "Blue") +
    theme_classic() + 
    theme(
      # plot.title = element_markdown(lineheight = .75, size= 9, face="bold", hjust = 0.5),
      plot.title = element_text(size= 12, face = "bold", color = 'black', hjust = 0.5),
      axis.title = element_text(size = 9, face="bold"),
      legend.title = element_blank(), legend.position = "top",
      panel.border = element_rect(color = "black", fill = NA, linewidth = 1)
    )


library(ggpubr)

# mh_it1<- ggarrange(iteration_plot1, zoomed_plot1,  
#                              common.legend = TRUE, nrow = 2, ncol = 1)

# pmh_it1<- annotate_figure( mh_it1, top = text_grob(expression(paste0("50000 iteration using M-H algorithm to sample from ", gamma(2,4))), color = "#0d233a", face = "bold", size = 14))

mh_d1 <- ggarrange(iteration_plot1, zoomed_plot1, prop_plot1, plot_mh1,  
                             common.legend = FALSE, nrow = 4, ncol = 1)

# pmh_d1<- annotate_figure( mh_d1, top = text_grob("Simulated draws from gamma(2,4) using M-H algorithm", color = "#0d233a", face = "bold", size = 10))

```

```{r include=TRUE}

print(mh_d1)


```

\newpage
* Approach 2: Normal distribution as a proposal with variance = 2 and $\theta_0 = 5$

```{r include=FALSE, echo=FALSE}
mh_df2 <- data.frame(theta = approach_2$theta) %>% 
  mutate(index = seq_along(1: nrow(.)),
         burn_in = if_else(index <= 10000, "Burn in", "Accepted"))

prop_plot2 <- ggplot() + 
  geom_density(aes( x = rgamma(40000, 2, 4)), col = "#003153", lwd = 1) + 
  geom_density(aes( x = rnorm(40000, 5, sqrt(2))), col = "red3", lwd = 1) +
  
  labs(title = expression(paste("Actual density ", Gamma(2, 4), " vs proposal ", N(theta = 5, 2.0))),
       x = expression(theta),
       y =  "Density") +
  theme_classic() + 
  theme(
    # plot.title = element_markdown(lineheight = .75, size= 9, face="bold", hjust = 0.5),
    plot.title = element_text(size= 12, face = "bold", color = 'black', hjust = 0.5),
    axis.title = element_text(size = 9, face="bold"),
    legend.title = element_blank(),
    legend.text = element_text(size = 9, face = "bold"), legend.position = c(.95, .9),
    legend.background = element_rect(color = "black", fill= NA, size = 1, linetype= "solid"),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1)) +
  scale_color_manual(values = col_pal)



iteration_plot2 <- ggplot(data = mh_df2) + 
  geom_line(aes(x = index, y = theta, color = burn_in)) +
  labs(title = "Metropolis Hastings using Normal proposal with starting point 5",
       x = "Iteration Number",
       y = expression(theta)) +
  theme_classic() + 
  theme(
    # plot.title = element_markdown(lineheight = .75, size= 9, face="bold", hjust = 0.5),
    plot.title = element_text(size= 12, face = "bold", color = 'black', hjust = 0.5),
    axis.title = element_text(size = 9, face="bold"),
    legend.title = element_blank(),
    legend.text = element_text(size = 9, face = "bold"), legend.position = c(.95, .9),
    legend.background = element_rect(color = "black", fill= NA, size = 1, linetype= "solid"),
    panel.border = element_rect(color = "black", fill = NA, size = 1)) +
  scale_color_manual(values = col_pal)

zoomed_plot2 <-  mh_df2 %>% 
  filter(9950< index & index <10050) %>% 
  ggplot(data = .) + 
  geom_line(aes(x = index, y = theta, color = burn_in)) +
  labs(title = "100 iteration Zoomed in ",
       x = "Iteration Number",
       y = expression(theta)) +
  theme_classic() + 
  theme(
    # plot.title = element_markdown(lineheight = .75, size= 9, face="bold", hjust = 0.5),
    plot.title = element_text(size= 12, face = "bold", color = 'black', hjust = 0.5),
    axis.title = element_text(size = 9, face="bold"),
    legend.title = element_blank(),
    legend.text = element_text(size = 9, face = "bold"), legend.position = "op",
    legend.background = element_rect(color = "black", fill= NA, size = 1, linetype= "solid"),
    panel.border = element_rect(color = "black", fill = NA, size = 1)) +
  scale_color_manual(values = col_pal)


plot_mh2 <- ggplot(data = filter(mh_df2, burn_in == "Accepted")) +
  geom_histogram(aes(x = theta, y= after_stat(density)), bins = 100, fill = "white", col = "#003153") + 
  geom_density(aes( x = rgamma(40000, 2, 4)), col = "seagreen", lwd = 1) +
  labs(title = expression(paste("Histogram of simulated sample using M-H algorithm from ", Gamma(2, 4)), " density"),
       x = expression(theta),
       y = "Density", color = "Blue") +
  theme_classic() + 
  theme(
    # plot.title = element_markdown(lineheight = .75, size= 9, face="bold", hjust = 0.5),
    plot.title = element_text(size= 12, face = "bold", color = 'black', hjust = 0.5),
    axis.title = element_text(size = 9, face="bold"),
    legend.title = element_blank(), legend.position = "top",
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1)
  )


library(ggpubr)

# mh_it2<- ggarrange(iteration_plot2, zoomed_plot2,  
#                    common.legend = TRUE, nrow = 2, ncol = 1)
# 
# pmh_it2<- annotate_figure( mh_it2, top = text_grob(expression(paste0("50000 iteration using M-H algorithm to sample from ", gamma(2,4))), color = "#0d233a", face = "bold", size = 14))

mh_d2 <- ggarrange(iteration_plot2, zoomed_plot2, prop_plot2, plot_mh2,  
                   common.legend = FALSE, nrow = 4, ncol = 1)

# pmh_d2<- annotate_figure( mh_d2, top = text_grob("Simulated draws from gamma(2,4) using M-H algorithm", color = "#0d233a", face = "bold", size = 10))
```

```{r include=TRUE}

# print(mh_it2)
print(mh_d2)

```

\newpage

* Approach 2: Normal distribution as a proposal with variance = 2 and $\theta_0 = 5$
```{r include = TRUE}

cat("\tAcceptance Rate:", approach_2$acceptance_rate, 
    "\n\tEstimated Mean:", estimated_mean_2, 
    "\n\tDifference from True Mean:", abs(estimated_mean_2 - true_mean))
```
## Solution 5:
Metropolis-Hastings Simulation to draw samples from density:
$$\theta \sim Gamma(2,\space 4)$$
```{r include=FALSE, echo=FALSE}

# mh_normal_df 
#  Q5: lognormal proposal
alpha <- 2; beta <- 4
target_density <- function(x){
  return(dgamma(x, shape = alpha, rate = beta))
}

true_mean <- alpha/beta
mh_alg_lnorm <- function( initial_theta = 5, num_steps, burn_in, proposal_variance){
  
  set.seed(729)
  theta <- numeric(num_steps)
  theta[1] <- initial_theta
  
  accepted_proposal <- 0
  
  for(i in 2 : num_steps){
    # propose a theta value from Normal density proposal
    candidate <- rlnorm(1, meanlog = log(theta[i-1]), sdlog = sqrt(proposal_variance))
    
    # Target ratio: decides whether we accept the proposed theta as current
    acceptance_ratio <- min(1, (target_density(candidate)/target_density(theta[i-1]))*(candidate/theta[i-1])) # markov kernel
    
    unif_value <- runif(1)
    
    if(unif_value < acceptance_ratio){
      
      # accept proposed theta
      theta[i] <- candidate
      accepted_proposal <- accepted_proposal + 1
      
    } else {
      
      # reject proposal and take immediate previous theta as current theta
      theta[i] <- theta[i - 1]
    }
    
  }
  
  theta_aburn <- theta[(burn_in + 1):num_steps] # samples after burn_in
  
  # require to decide proposal variance
  acceptance_rate <- accepted_proposal/num_steps
  
    return(list(theta = theta, theta_aburn = theta_aburn, acceptance_rate = acceptance_rate))
}

#  using proposal variance = 0.1
approach_i <- mh_alg_lnorm(num_steps = 50000, burn_in = 10000, proposal_variance = 0.25)

#  using proposal variance = 2
approach_ii <- mh_alg_lnorm(num_steps = 50000, burn_in = 10000, proposal_variance = 25)

# Calculate the estimated mean for approach 1
estimated_mean_i <- mean(approach_i$theta)

# Calculate the estimated mean for approach 2
estimated_mean_ii <- mean(approach_ii$theta)

```
From target distribution True mean is 0.5

* Approach 1: Log-Normal distribution as a proposal with variance = 0.25 and $\theta_0 = 5$

```{r include = TRUE}

# Print the results
cat("\tAcceptance Rate:", approach_i$acceptance_rate, 
    "\n\tEstimated Mean:", estimated_mean_i, 
    "\n\tDifference from True Mean:", abs(estimated_mean_i - true_mean))

```



\newpage
* Approach 1: Log-Normal distribution as a proposal with variance = 0.25 and $\theta_0 = 5$

```{r include=FALSE, echo=FALSE}

mh_df1 <- data.frame(theta = approach_i$theta) %>% 
  mutate(index = seq_along(1: nrow(.)),
         burn_in = if_else(index <= 10000, "Burn in", "Accepted"))
col_pal <- c("#003153", "red3")

prop_plot1 <- ggplot() + 
    geom_density(aes( x = rgamma(40000, 2, 4)), col = "#003153", lwd = 1) +
    geom_density(aes( x = rlnorm(40000, 5, sqrt(0.25))), col = "red3", lwd = 1) + 
    xlim(0, 50) +
    
    labs(title = expression(paste("Actual density ", Gamma(2, 4), " vs proposal ", Log-Normal(theta = 5, 0.25))),
       x = expression(theta),
       y =  "Density") +
    theme_classic() + 
    theme(
      # plot.title = element_markdown(lineheight = .75, size= 9, face="bold", hjust = 0.5),
      plot.title = element_text(size= 12, face = "bold", color = 'black', hjust = 0.5),
      axis.title = element_text(size = 9, face="bold"),
      legend.title = element_blank(),
      legend.text = element_text(size = 9, face = "bold"), legend.position = c(.95, .9),
      legend.background = element_rect(color = "black", fill= NA, linewidth = 1, linetype= "solid"),
      panel.border = element_rect(color = "black", fill = NA, linewidth = 1)) +
      scale_color_manual(values = col_pal)



iteration_plot1 <- ggplot(data = mh_df1) + 
    geom_line(aes(x = index, y = theta, color = burn_in)) +
    labs(title = "Metropolis Hastings using Normal proposal with starting point 5",
         x = "Iteration Number",
         y = expression(theta)) +
    theme_classic() + 
    theme(
      # plot.title = element_markdown(lineheight = .75, size= 9, face="bold", hjust = 0.5),
      plot.title = element_text(size= 12, face = "bold", color = 'black', hjust = 0.5),
      axis.title = element_text(size = 9, face="bold"),
      legend.title = element_blank(),
      legend.text = element_text(size = 9, face = "bold"), legend.position = c(.95, .9),
      legend.background = element_rect(color = "black", fill= NA, size = 1, linetype= "solid"),
      panel.border = element_rect(color = "black", fill = NA, size = 1)) +
    scale_color_manual(values = col_pal)

zoomed_plot1 <-  mh_df1 %>% 
  filter(9950< index & index <10050) %>% 
  ggplot(data = .) + 
  geom_line(aes(x = index, y = theta, color = burn_in)) +
  labs(title = "100 iteration Zoomed in ",
       x = "Iteration Number",
       y = expression(theta)) +
  theme_classic() + 
    theme(
      # plot.title = element_markdown(lineheight = .75, size= 9, face="bold", hjust = 0.5),
      plot.title = element_text(size= 12, face = "bold", color = 'black', hjust = 0.5),
      axis.title = element_text(size = 9, face="bold"),
      legend.title = element_blank(),
      legend.text = element_text(size = 9, face = "bold"), legend.position = c(.95, .9),
      legend.background = element_rect(color = "black", fill= NA, size = 1, linetype= "solid"),
      panel.border = element_rect(color = "black", fill = NA, size = 1)) +
    scale_color_manual(values = col_pal)
  

plot_mh1 <- ggplot(data = filter(mh_df1, burn_in == "Accepted")) +
    geom_histogram(aes(x = theta, y= after_stat(density)), bins = 100, fill = "white", col = "#003153") + 
    geom_density(aes( x = rgamma(40000, 2, 4)), col = "seagreen", lwd = 1) +
    labs(title = expression(paste("Histogram of simulated sample using M-H algorithm from ", Gamma(2, 4)), " density"),
         x = expression(theta),
         y = "Density", color = "Blue") +
    theme_classic() + 
    theme(
      # plot.title = element_markdown(lineheight = .75, size= 9, face="bold", hjust = 0.5),
      plot.title = element_text(size= 12, face = "bold", color = 'black', hjust = 0.5),
      axis.title = element_text(size = 9, face="bold"),
      legend.title = element_blank(), legend.position = "top",
      panel.border = element_rect(color = "black", fill = NA, linewidth = 1)
    )


library(ggpubr)

# mh_it1<- ggarrange(iteration_plot1, zoomed_plot1,  
#                              common.legend = TRUE, nrow = 2, ncol = 1)

# pmh_it1<- annotate_figure( mh_it1, top = text_grob(expression(paste0("50000 iteration using M-H algorithm to sample from ", gamma(2,4))), color = "#0d233a", face = "bold", size = 14))

mh_d1 <- ggarrange(iteration_plot1, zoomed_plot1, prop_plot1, plot_mh1,  
                             common.legend = FALSE, nrow = 4, ncol = 1)

# pmh_d1<- annotate_figure( mh_d1, top = text_grob("Simulated draws from gamma(2,4) using M-H algorithm", color = "#0d233a", face = "bold", size = 10))

```

```{r include=TRUE}

print(mh_d1)


```

\newpage
* Approach 2: Log Normal distribution as a proposal with variance = 25 and $\theta_0 = 5$

```{r include=FALSE, echo=FALSE}
mh_df2 <- data.frame(theta = approach_ii$theta) %>% 
  mutate(index = seq_along(1: nrow(.)),
         burn_in = if_else(index <= 10000, "Burn in", "Accepted"))

prop_plot2 <- ggplot() + 
  geom_density(aes( x = rgamma(40000, 2, 4)), col = "#003153", lwd = 1) +
  geom_density(aes( x = rlnorm(40000, 5, sqrt(25))), col = "red3", lwd = 1) + 
  xlim(0, 50) +
  
  labs(title = expression(paste("Actual density ", Gamma(2, 4), " vs proposal ", LogNormal(theta = 5, 25.0))),
       x = expression(theta),
       y =  "Density") +
  theme_classic() + 
  theme(
    # plot.title = element_markdown(lineheight = .75, size= 9, face="bold", hjust = 0.5),
    plot.title = element_text(size= 12, face = "bold", color = 'black', hjust = 0.5),
    axis.title = element_text(size = 9, face="bold"),
    legend.title = element_blank(),
    legend.text = element_text(size = 9, face = "bold"), legend.position = c(.95, .9),
    legend.background = element_rect(color = "black", fill= NA, size = 1, linetype= "solid"),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1)) +
  scale_color_manual(values = col_pal)



iteration_plot2 <- ggplot(data = mh_df2) + 
  geom_line(aes(x = index, y = theta, color = burn_in)) +
  labs(title = "Metropolis Hastings using Log Normal proposal with starting point 5",
       x = "Iteration Number",
       y = expression(theta)) +
  theme_classic() + 
  theme(
    # plot.title = element_markdown(lineheight = .75, size= 9, face="bold", hjust = 0.5),
    plot.title = element_text(size= 12, face = "bold", color = 'black', hjust = 0.5),
    axis.title = element_text(size = 9, face="bold"),
    legend.title = element_blank(),
    legend.text = element_text(size = 9, face = "bold"), legend.position = c(.95, .9),
    legend.background = element_rect(color = "black", fill= NA, size = 1, linetype= "solid"),
    panel.border = element_rect(color = "black", fill = NA, size = 1)) +
  scale_color_manual(values = col_pal)

zoomed_plot2 <-  mh_df2 %>% 
  filter(9950< index & index <10050) %>% 
  ggplot(data = .) + 
  geom_line(aes(x = index, y = theta, color = burn_in)) +
  labs(title = "100 iteration Zoomed in ",
       x = "Iteration Number",
       y = expression(theta)) +
  theme_classic() + 
  theme(
    # plot.title = element_markdown(lineheight = .75, size= 9, face="bold", hjust = 0.5),
    plot.title = element_text(size= 12, face = "bold", color = 'black', hjust = 0.5),
    axis.title = element_text(size = 9, face="bold"),
    legend.title = element_blank(),
    legend.text = element_text(size = 9, face = "bold"), legend.position = "op",
    legend.background = element_rect(color = "black", fill= NA, size = 1, linetype= "solid"),
    panel.border = element_rect(color = "black", fill = NA, size = 1)) +
  scale_color_manual(values = col_pal)


plot_mh2 <- ggplot(data = filter(mh_df2, burn_in == "Accepted")) +
  geom_histogram(aes(x = theta, y= after_stat(density)), bins = 100, fill = "white", col = "#003153") + 
  geom_density(aes( x = rgamma(40000, 2, 4)), col = "seagreen", lwd = 1) +
  labs(title = expression(paste("Histogram of simulated sample using M-H algorithm from ", Gamma(2, 4)), " density"),
       x = expression(theta),
       y = "Density", color = "Blue") +
  theme_classic() + 
  theme(
    # plot.title = element_markdown(lineheight = .75, size= 9, face="bold", hjust = 0.5),
    plot.title = element_text(size= 12, face = "bold", color = 'black', hjust = 0.5),
    axis.title = element_text(size = 9, face="bold"),
    legend.title = element_blank(), legend.position = "top",
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1)
  )


library(ggpubr)

# mh_it2<- ggarrange(iteration_plot2, zoomed_plot2,  
#                    common.legend = TRUE, nrow = 2, ncol = 1)
# 
# pmh_it2<- annotate_figure( mh_it2, top = text_grob(expression(paste0("50000 iteration using M-H algorithm to sample from ", gamma(2,4))), color = "#0d233a", face = "bold", size = 14))

mh_d2 <- ggarrange(iteration_plot2, zoomed_plot2, prop_plot2, plot_mh2,  
                   common.legend = FALSE, nrow = 4, ncol = 1)

# pmh_d2<- annotate_figure( mh_d2, top = text_grob("Simulated draws from gamma(2,4) using M-H algorithm", color = "#0d233a", face = "bold", size = 10))
```

```{r include=TRUE}

# print(mh_it2)
print(mh_d2)

```

\newpage

* Approach 2: Log Normal distribution as a proposal with variance = 25 and $\theta_0 = 5$
```{r include = TRUE}


cat("\tAcceptance Rate:", approach_ii$acceptance_rate, 
    "\n\tEstimated Mean:", estimated_mean_ii, 
    "\n\tDifference from True Mean:", abs(estimated_mean_ii - true_mean))
```

## solution 6:
i) $$f(\theta) = c * \theta^{3}$$ to be a valid PDF, c need to be 4

ii) Simulation using Inverse CDDF method:

|     Since, $$F_\theta(t) = t^4$$
|     Thus, $$\theta = F^{-1}(u) = u^{\frac{1}{4}}$$ 
|     where, $$U \sim unif(0,1)$$

```{r include=FALSE, echo=FALSE}
# -------------- Question 6: Inverse CDF method for a expression:
# i) c = 4

# ii) inverse cdf of f(x) = 4 * x^3 is u ^(1/4)

# Desired3 number of observations
n <- 10000

# Generate observations using inverse CDF method
inverse_cdf <- function(n) {
  
  # Generate uniform random numbers
  u <- runif(n) # default unif(0, 1)
  
  # Apply inverse CDF transformation
  observations <- u^(1/4)
  return(observations)
}

icdf_obs <- inverse_cdf(n)

# hist(icdf_obs, probability = TRUE, col = "grey", 
#      main=expression(paste("Histogram of simulated sample from ", f(x), "=", 4*x^3)))
# curve(4 * x^3, 0,25, n=1000,col='red3',lwd=1,add=TRUE)

theta = seq(0, 1, length.out = 10000)
ftheta <- 4*theta^3
inv_df <- data.frame(theta = theta, ftheta = ftheta, icdf_obs = icdf_obs)

plot_icdf <- ggplot(data = inv_df) +
  geom_histogram(aes(x = icdf_obs, y= after_stat(density)), bins = 100, fill = "white", col = "#003153") + 
  geom_line(aes(x = theta, y = ftheta), col = "seagreen", lwd = 1) +
  xlim(0, 1) +
  labs(title = expression(paste("Histogram of simulated draws using inverse CDF method for ", f(x), " = ", 4*x^3)),
         x = expression(theta),
         y = "Density") +
  theme_classic() + 
  theme(
    # plot.title = element_markdown(lineheight = .75, size= 9, face="bold", hjust = 0.5),
    plot.title = element_text(size = 12, face = "bold", color = 'black', hjust = 0.5),
    axis.title = element_text(size = 9, face="bold"),
    legend.title = element_blank(), legend.position = "top",
    panel.border = element_rect(color = "black", fill = NA, size = 1)
    )
```

```{r include=TRUE, fig.width=12, fig.height=8}
print(plot_icdf)
```

\newpage

iii) Simulation usin Rejection sampling:
```{r, include=FALSE, echo=FALSE}

# iii) Rejection sampling

# Desired3 the number of observations
n <- 10000

# Generate observations using rejection sampling
rej_sampling <- function(n) {
  
  observations <- numeric(n)
  accepted <- 0
  rejected <- 0
  
  while (accepted < n) {
    x <- runif(1)  # Generate a random number in the range [0, 1]
    y <- runif(1)  # Generate a random number in the range [0, 1]
    
    if (y <  x^3) { # given PDF
      observations[accepted + 1] <- x
      accepted <- accepted + 1
    } else {
      rejected <- rejected + 1
    }
  }
  
  return(list(samples = observations, rejection_number = rejected))
}

# Generate observations using rejection sampling
rej_samples <- rej_sampling(n)

theta = seq(0,1, length.out = 10000)
ftheta <- 4*theta^3
rej_df <- data.frame(theta = theta, ftheta = ftheta, rej_sample = rej_samples$samples)

plot_rej <- ggplot(data = rej_df) +
    geom_histogram(aes(x = rej_sample, y= after_stat(density)), bins = 100, fill = "white", col = "#003153") + 
    geom_line(aes(x = theta, y = ftheta), col = "seagreen", lwd = 1) +
    xlim(0, 1) +
    labs(title = expression(paste("Histogram of accepted draws using rejection sampling for ", f(x), " = ", 4*x^3)),
         x = expression(theta),
         y = "Density", color = "black") +
    theme_classic() + 
    theme(
      # plot.title = element_markdown(lineheight = .75, size= 9, face="bold", hjust = 0.5),
      plot.title = element_text(size= 14, face = "bold", color = 'black', hjust = 0.5),
      axis.title = element_text(size = 9, face="bold"),
      legend.title = element_blank(), legend.position = "top",
      panel.border = element_rect(color = "black", fill = NA, size = 1)
    )
```

```{r include=TRUE, fig.width=12, fig.height= 8}
print(plot_rej)
```

```{r include=FALSE, echo=FALSE}

# (iv) the average number of trials per accepted observation in Part (iii).
average_trials<- (rej_samples$rejection_number + length(rej_samples$samples)) / length(rej_samples$samples)

# v) Compare the two sets of observations generated in Part (ii) and Part (iii) in terms of
# empirical mean, variance and 60% quantile.

# Calculate empirical mean, variance, and 60% quantile for inverse CDF method
mean_icdf <- mean(icdf_obs)
variance_icdf <- var(icdf_obs)
quantile_icdf <- quantile(icdf_obs, 0.6)

# Calculate empirical mean, variance, and 60% quantile for rejection sampling
mean_rejection_sampling <- mean(rej_samples$samples)
variance_rejection_sampling <- var(rej_samples$samples)
quantile_rejection_sampling <- quantile(rej_samples$samples, 0.6)
```

iv) Summary results from two sets of observations 

* Using Inverse CDF:

```{r include=TRUE}
# Print the results
cat("\tEmpirical Mean:", mean_icdf, 
    "\n\tVariance:", variance_icdf, 
    "\n\t60% Quantile:", quantile_icdf, "\n")
```

* Using Rejection sampling:

```{r include = TRUE}

cat("\tEmpirical Mean:", mean_rejection_sampling, 
    "\n\tVariance:", variance_rejection_sampling, 
    "\n\t60% Quantile:", quantile_rejection_sampling, "\n")
```





